---
title: "Data_exploration"
author: "Claudia Bardales"
date: "`r Sys.Date()`"
output: word_document
---

```{r }

my_data <- read.csv("my_data_clean_fl.csv")
head(my_data)
colnames(my_data)
```
#libraries
```{r }
library(tidyverse) 
library(cowplot)
#Maps
library(tigris)
library(sf)
library(scales) 
library(plotly)

#radar
#install.packages("devtools")
devtools::install_github("ricardo-bion/ggradar")
# install.packages(c("systemfonts", "ragg", "pkgbuild", "fansi"))
library(ggradar)
library(tibble)

library(stringr)

```
#check total claims and total drug cost per Zipcode
```{r }

grp1 <- my_data |>
  select(Prscrbr_City, Prscrbr_zip5,Tot_Clms, Tot_Drug_Cst) |>
  group_by(Prscrbr_City, Prscrbr_zip5) |>
  summarise(
    Total_Claims_K= sum(Tot_Clms/1000, na.rm = TRUE),
    Total_Drug_Cost_MM = round(sum(Tot_Drug_Cst, na.rm = TRUE) / 1000000, 2) 
  , .groups = "drop")  |>
  arrange(desc(Total_Claims_K))

```

#Convert zipcode to text
```{r }
grp1$Prscrbr_zip5 <- as.character(grp1$Prscrbr_zip5)
```

```{r }
grp1

```

```{r }
fl_zip <- zctas(state = "FL", year = 2010, class = "sf")

fl_zip <- left_join(fl_zip, grp1, by = c("ZCTA5CE10" = "Prscrbr_zip5"))

#Total Claims maps per Zipcode
ggplot(data = fl_zip) +
  geom_sf(aes(fill = Total_Claims_K)) +
  scale_fill_viridis_c() +  # color for better visualization
  labs(title = "Total Claims by Zip Code (Florida) in thousands", fill = "Total Claims", 
       subtitle = "Source: CMS 2022") +
  theme_minimal()+
    theme(
        plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 10, color = "gray50"),
        panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
        ) 

#Total Total_Drug_Cost_MM maps per Zipcode
ggplot(data = fl_zip) +
  geom_sf(aes(fill = Total_Drug_Cost_MM)) +
  scale_fill_viridis_c() +  # Color for better visualization
  labs(title = "Total Drug Cost by Zip Code (Florida) in millons", fill = "Total Drug Cost",
              subtitle = "Source: CMS 2022") +
  theme_minimal()+
    theme(
        plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 10, color = "gray50"),
        panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
        ) 

```




```{r }

grp2 <- my_data |>
  select(Prscrbr_City,Tot_Clms, Tot_Drug_Cst) |>
  group_by(Prscrbr_City) |>
  summarise(
    Total_Claims_K= sum(Tot_Clms/1000, na.rm = TRUE),
    Total_Drug_Cost_MM = round(sum(Tot_Drug_Cst, na.rm = TRUE) / 1000000, 2) 
  , .groups = "drop")  |>
  arrange(desc(Total_Claims_K))


grp2
#total cities in Fl that submitted claims in 2022 is 644
```

#top 10 cities that made claims 
```{r }

# Cities of interest
cities_of_interest <- c("Miami", "Jacksonville", "Tampa", "Orlando", "Hialeah",
                        "Fort Myers", "Sarasota", "Ocala", "Lakeland", "Naples")

filtered_data <- my_data |>
  filter(Prscrbr_City %in% cities_of_interest)


filtered_data

```
#A chart that shows the distribution of a single categorical variable
```{r }

#total claims by city

filtered_data_by_city<-filtered_data|>
  select(Prscrbr_City,Tot_Clms)|>
  group_by(Prscrbr_City)|>
  summarise(total_claims_per_city=sum(Tot_Clms)/1000)|>
  arrange(desc(total_claims_per_city))
filtered_data_by_city

min_val <- min(filtered_data_by_city$total_claims_per_city, na.rm = TRUE)
max_val <- max(filtered_data_by_city$total_claims_per_city, na.rm = TRUE)

a <- ggplot(data = filtered_data_by_city, 
            aes(x = reorder(Prscrbr_City, -total_claims_per_city), 
                y = total_claims_per_city, 
                fill = Prscrbr_City)) +
  geom_bar(stat = "identity", color = "black") + 
  #add labels
  geom_text(aes(label = label_number(accuracy = 0.01, big.mark = "'")(total_claims_per_city)), 
            vjust = -0.5, size = 3) +
  labs(title = "Top 10 Cities in Florida by Total Claims",
       subtitle = "Source: CMS 2022",
       x = "City", 
       y = "Total Claims (in thousands)") +
  #theme
  theme_minimal() + 
  theme(
        plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 10, color = "gray50"),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()
        ) +
  guides(fill = "none") 

a

 

```

#A chart that shows the distribution of a single quantitative variable
```{r }
filtered_data_by_city_Cost<-filtered_data|>
  select(Prscrbr_City, Tot_Drug_Cst)
filtered_data_by_city_Cost

min_a<-min(filtered_data_by_city_Cost$Tot_Drug_Cst)
min_a

max_a<-max(filtered_data_by_city_Cost$Tot_Drug_Cst)
max_a
summary(filtered_data_by_city_Cost$Tot_Drug_Cst)

#outlier
Q1 <- quantile(filtered_data_by_city_Cost$Tot_Drug_Cst, 0.25, na.rm = TRUE)
Q3 <- quantile(filtered_data_by_city_Cost$Tot_Drug_Cst, 0.75, na.rm = TRUE)

IQR_value <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

# Count outliers
left_outliers <- sum(filtered_data_by_city_Cost$Tot_Drug_Cst < lower_bound, na.rm = TRUE)
right_outliers <- sum(filtered_data_by_city_Cost$Tot_Drug_Cst > upper_bound, na.rm = TRUE)
left_outliers
right_outliers
#There is no outlier to the left, there are 4843 outiler to the right




```

```{r }

dens <- ggplot(filtered_data_by_city_Cost, aes(x = Tot_Drug_Cst)) +
  geom_density(fill = "blue", alpha = 0.5) +
  scale_x_continuous(
    #limits = c(0, 1e6),  
    labels = label_number(scale = 1e-6, big.mark = "'", suffix = " MM")  # Format labels
  ) +
  labs(
    title = "Density Plot of Total Drug Cost",
    x = NULL, 
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.ticks.x = element_line(),   # Ensure tick marks are visible
    axis.text.x = element_text(color = "black"),  # Keep tick labels visible
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )



```

```{r }

box<-ggplot(filtered_data_by_city_Cost, aes(x = Tot_Drug_Cst)) +
  geom_boxplot(fill = "skyblue", color = "black", alpha = 0.1, outlier.color = "purple4")  +
  labs(
       x = "Total Drug Cost (in MM $)") +
  theme_minimal()+
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),    
    axis.line.x = element_blank())


```

```{r}
#join files
plot_grid(dens,box, ncol=1, rel_heights = c(0.80,0.20) , align='v', axis='lr')

```

#A chart that shows the distribution of two categorical variables
```{r }


cat_cat <- filtered_data |>
  select(Prscrbr_City, Brnd_Tot_Clms, Gnrc_Tot_Clms, Tot_Clms) |>
  group_by(Prscrbr_City) |>
  summarise(
    Brand = sum(Brnd_Tot_Clms, na.rm = TRUE),
    Generic = sum(Gnrc_Tot_Clms, na.rm = TRUE),
    Total = sum(Tot_Clms, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(Other = Total - Brand - Generic)|>
  pivot_longer(cols=c(Brand, Generic, Other),
              names_to = "Claim_Type",
              values_to = "Total_Claims")

cat_cat<- cat_cat|>
  select(Prscrbr_City,Claim_Type,Total_Claims)

#maintain order:
cat_cat$Claim_Type <- factor(cat_cat$Claim_Type, levels = c( "Other", "Brand", "Generic"))

#porcentage
cat_cat_percent <- cat_cat |>
  group_by(Prscrbr_City) |>
  mutate(
    Percent_Claims = Total_Claims / sum(Total_Claims) * 100
  ) |>
  ungroup()
```

```{r }
b <- ggplot(cat_cat_percent, 
            aes(x = reorder(Prscrbr_City, -Total_Claims), 
                y = Percent_Claims, 
                fill = Claim_Type)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.45) +
  geom_text(aes(label = paste0(round(Percent_Claims, 1), "%")),
            position = position_stack(vjust = 0.5),
            size = 3, color = "black") +
  labs(
    title = "Claim Type Composition by City (as % of Total Claims)",
    subtitle = "Source: CMS 2022",
    x = "City",
    y = "Percentage of Total Claims",
    fill = "Claim Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )

b
```


```{r }
cat_cat_thousands <- cat_cat |>
  mutate(Total_Claims = Total_Claims / 1000)

max_val <- 11000  
y_breaks <- pretty(c(0, max_val), n = 10)

c <- ggplot(cat_cat_thousands, 
            aes(x = reorder(Prscrbr_City, -Total_Claims), 
                y = Total_Claims, 
                fill = Claim_Type)) +
  geom_bar(stat = "identity", color = "black") +

  labs(
    title = "Total Claims Composition by City",
    subtitle = "Source: CMS 2022",
 #   x = "City",
    y = "Total Claims (in thousands)",
    fill = "Claim Type"
  ) +
  
  scale_y_continuous(
    breaks = y_breaks,
    limits = c(0, max(y_breaks)),
    labels = label_number(accuracy = 1, big.mark = "'")
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_line(),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )

c
```
```{r }
d <- ggplot(cat_cat_percent, 
            aes(x = reorder(Prscrbr_City, -Total_Claims), 
                y = Percent_Claims, 
                fill = Claim_Type)) +
  geom_bar(stat = "identity", color = "black") +
  labs(x = "City",
       y= "%")+
  geom_text(
    data = cat_cat_percent |> filter(Claim_Type %in% c("Generic")),
    aes(label = paste0(round(Percent_Claims, 1), "%")),
    position = position_stack(vjust = 0.4), 
    size = 3, color = "black"
  ) +

  theme_minimal() +
  theme(
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),
    #axis.text.x = element_blank(),
    #axis.title.x = element_blank(),  # remove x-axis title
    #axis.title.y = element_blank(),
    legend.position = "none"
  )
d

```

```{r }

plot_grid(c,d, ncol=1, rel_heights = c(0.80,0.20) , align='v', axis='lr')
```


```{r }

```

```{r }


```
#A chart that shows the relationship between two quantitative variables
#each obs refer to Prscrbr_NPI, here we are looking in average how much prescription and mean claim each doctor do anually
```{r }
city_summary_data <- filtered_data |>
  select(Prscrbr_NPI, Prscrbr_City, Tot_Drug_Cst, Tot_Clms) |>
  group_by(Prscrbr_City) |>
  summarise(
    Drug_Cost_mean = mean(Tot_Drug_Cst, na.rm = TRUE),
    Total_Claims_mean = mean(Tot_Clms, na.rm = TRUE),
    Number_Prescribers=n(),
    .groups = 'drop')
city_summary_data

```


```{r }
x_max<-max(city_summary_data$Total_Claims_mean)
x_min<-min(city_summary_data$Total_Claims_mean)
y_max<-max(city_summary_data$Drug_Cost_mean)
y_min<-min(city_summary_data$Drug_Cost_mean)

lgd_min<-min(city_summary_data$Number_Prescribers)
lgd_max<-max(city_summary_data$Number_Prescribers)

ggplot(city_summary_data, aes(
  x = Total_Claims_mean,
  y = Drug_Cost_mean,
  size = Number_Prescribers,
  label = Prscrbr_City
)) +
  geom_point(alpha = 0.5, color = "blue", stroke = 0.5) +
  geom_text(vjust = -1, size = 2.5, hjust=-0.001) +
  scale_size(range = c(3, 12)) +
  labs(
    title = "City Comparison: Drug Cost vs. Claim Volume",
    subtitle = "Source: CMS 2022",
    x = "Average Claims per Prescriber",
    y = "Average Drug Cost per Prescriber",
    size = "Number of Prescribers"
  ) +
  scale_x_continuous(
    limits = c(x_min-50, x_max + 100),
    breaks = seq(x_min, x_max + 50, length.out = 5),
    labels = label_number(big.mark = "'")
  ) +
  scale_y_continuous(
    limits = c(y_min-50, y_max + 50),
    breaks = seq(y_min, y_max + 50, length.out = 5),
    labels = label_number(big.mark = "'")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(),
    axis.text = element_text(size = 9),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black")
  )

```

```{r }
eff_per_city <- filtered_data |>
  select(Prscrbr_NPI, Prscrbr_City, Tot_Drug_Cst, Tot_Clms)
eff_per_city
```

#pending- Reduce limits , increase to see values under 10k 

```{r }

ordered_cities <- filtered_data_by_city$Prscrbr_City

eff_per_city <- filtered_data |>
  select(Prscrbr_NPI, Prscrbr_City, Tot_Clms) |>
  mutate(Prscrbr_City = factor(Prscrbr_City, levels = ordered_cities))


ggplot(eff_per_city, aes(x = Prscrbr_City, y = Tot_Clms, fill = Prscrbr_City)) +
  geom_boxplot(color = "black", alpha = 0.3, outlier.color = "red") +
  scale_y_continuous(
    breaks = pretty(eff_per_city$Tot_Clms, n = 8),
    labels = label_number(scale = 1e-3, suffix = "k", big.mark = "'")
  ) +
  labs(
    title = "Distribution of Total Claims per Prescriber by City",
    subtitle = "Source: CMS 2022",
    x = "City",
    y = "Total Claims"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    axis.text.x = element_text(size = 9, vjust = 0.5),
    axis.line = element_line(color = "black"),
    panel.grid = element_blank()
  ) +
  guides(fill = "none")  

```


```{r }


ordered_cities <- filtered_data_by_city$Prscrbr_City

eff_per_city_2 <- filtered_data |>
  select(Prscrbr_NPI, Prscrbr_City, Tot_Clms) |>
  mutate(Prscrbr_City = factor(Prscrbr_City, levels = ordered_cities))

eff_per_city_summary <- eff_per_city_2 |>
  group_by(Prscrbr_City) |>
  summarise(
    min = min(Tot_Clms, na.rm = TRUE),
    q1 = quantile(Tot_Clms, 0.25, na.rm = TRUE),
    q2 = quantile(Tot_Clms, 0.50, na.rm = TRUE),  # median
    q3 = quantile(Tot_Clms, 0.75, na.rm = TRUE),
    max = max(Tot_Clms, na.rm = TRUE),
    iqr = IQR(Tot_Clms, na.rm = TRUE),
    .groups = "drop"
  )

eff_per_city_summary

```

```{r }

ggplot(eff_per_city_2, aes(x = Prscrbr_City, y = Tot_Clms, fill = Prscrbr_City)) +
  geom_boxplot(color = "black", alpha = 0.3, outlier.color = "red") +
  scale_y_continuous(
    breaks = seq(0, 4000, by = 500),
    limits = c(0, 4000),
    labels = label_number(scale = 1e-3, suffix = "k", big.mark = "'")
  ) +
  labs(
    title = "Distribution of Total Claims per Prescriber by City",
    subtitle = "Source: CMS 2022",
    x = "City",
    y = "Total Claims"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    axis.text.x = element_text(size = 9, vjust = 0.5),
    axis.line = element_line(color = "black"),
    panel.grid = element_blank()
  ) +
  guides(fill = "none")

```
#map to check demographic of patients 
```{r }

demo<-filtered_data|>
  select(Prscrbr_City,
    Bene_Age_LT_65_Cnt,
    Bene_Age_65_74_Cnt,
    Bene_Age_75_84_Cnt,
    Bene_Age_GT_84_Cnt,
    Bene_Feml_Cnt,
    Bene_Male_Cnt,
    Bene_Race_Wht_Cnt,
    Bene_Race_Black_Cnt,
    Bene_Race_Api_Cnt,
    Bene_Race_Hspnc_Cnt,
    Bene_Race_Natind_Cnt,
    Bene_Race_Othr_Cnt
  )

demo_1<-demo|>
  group_by(Prscrbr_City)|>
  summarise(    sum(Bene_Age_LT_65_Cnt),
    sum(Bene_Age_65_74_Cnt),
    sum(Bene_Age_75_84_Cnt),
    sum(Bene_Age_GT_84_Cnt),
    sum(Bene_Feml_Cnt),
    sum(Bene_Male_Cnt),
    sum(Bene_Race_Wht_Cnt),
    sum(Bene_Race_Black_Cnt),
    sum(Bene_Race_Api_Cnt),
    sum(Bene_Race_Hspnc_Cnt),
    sum(Bene_Race_Natind_Cnt),
    sum(Bene_Race_Othr_Cnt)
  )

demo_2<-demo|>
  pivot_longer(
    cols = starts_with("Bene_Age"),  
    names_to = "Age_Group",         
    values_to = "Age_Count" 
  )

demo_2<-demo_2|>
  pivot_longer(
    cols = c("Bene_Feml_Cnt",
    "Bene_Male_Cnt"),  
    names_to = "Sex_Group",         
    values_to = "Sex_Count" 
  )

demo_2<-demo_2|>
  pivot_longer(
    cols = starts_with("Bene_Race"),  
    names_to = "Race_Group",         
    values_to = "Race_Count" 
  )
#write.csv(demo_1, "demo1.csv", row.names = FALSE)

```



```{r }

demo<-filtered_data|>
  select(Prscrbr_City,
    Bene_Age_LT_65_Cnt,
    Bene_Age_65_74_Cnt,
    Bene_Age_75_84_Cnt,
    Bene_Age_GT_84_Cnt,
    Bene_Feml_Cnt,
    Bene_Male_Cnt,
    Bene_Race_Wht_Cnt,
    Bene_Race_Black_Cnt,
    Bene_Race_Api_Cnt,
    Bene_Race_Hspnc_Cnt,
    Bene_Race_Natind_Cnt,
    Bene_Race_Othr_Cnt
  )


# Summarize the data with better names
demo_3 <- demo |>
  group_by(Prscrbr_City) %>%
  summarise(
    Age_LT_65_Cnt_Sum = sum(Bene_Age_LT_65_Cnt, na.rm = TRUE),
    Age_65_74_Cnt_Sum = sum(Bene_Age_65_74_Cnt, na.rm = TRUE),
    Age_75_84_Cnt_Sum = sum(Bene_Age_75_84_Cnt, na.rm = TRUE),
    Age_GT_84_Cnt_Sum = sum(Bene_Age_GT_84_Cnt, na.rm = TRUE),
    Female_Cnt_Sum = sum(Bene_Feml_Cnt, na.rm = TRUE),
    Male_Cnt_Sum = sum(Bene_Male_Cnt, na.rm = TRUE),
    White_Cnt_Sum = sum(Bene_Race_Wht_Cnt, na.rm = TRUE),
    Black_Cnt_Sum = sum(Bene_Race_Black_Cnt, na.rm = TRUE),
    Asian_Pacific_Islander_Cnt_Sum = sum(Bene_Race_Api_Cnt, na.rm = TRUE),
    Hispanic_Cnt_Sum = sum(Bene_Race_Hspnc_Cnt, na.rm = TRUE),
    Native_Indian_Cnt_Sum = sum(Bene_Race_Natind_Cnt, na.rm = TRUE),
    Other_Race_Cnt_Sum = sum(Bene_Race_Othr_Cnt, na.rm = TRUE))

demo_4 <- demo_3 |>
  pivot_longer(
    cols = -Prscrbr_City, 
    names_to = "Category", 
    values_to = "Count"     
  )

demo_4$Category <- recode(demo_4$Category,
                          "Age_LT_65_Cnt_Sum" = "Age < 65",
                          "Age_65_74_Cnt_Sum" = "Age 65-74",
                          "Age_75_84_Cnt_Sum" = "Age 75-84",
                          "Age_GT_84_Cnt_Sum" = "Age > 84",
                          "Female_Cnt_Sum" = "Female",
                          "Male_Cnt_Sum" = "Male",
                          "White_Cnt_Sum" = "White",
                          "Black_Cnt_Sum" = "Black",
                          "Asian_Pacific_Islander_Cnt_Sum" = "Asian",
                          "Hispanic_Cnt_Sum" = "Hispanic",
                          "Native_Indian_Cnt_Sum" = "Native Indian",
                          "Other_Race_Cnt_Sum" = "Other Race")


demo_4$Category <- str_wrap(demo_4$Category, width = 5)



demo_4$Category <- factor(demo_4$Category, 
                           levels = c(
                             "Age < 65", "Age 65-74", "Age 75-84", "Age > 84", 
                             "Female", "Male", 
                             "White", "Black",  "Hispanic", "Asian",
                             "Native Indian", "Other Race"
                           ))



```

```{r }
demo_4 <- demo_4 |>
  filter(!is.na(Category))

ggplot(demo_4, aes(x = Category, y = Prscrbr_City, fill = Count)) +
  geom_tile() +
  scale_fill_gradient(
    low = "white", high = "blue", 
    name = "Total Count", 
    labels = label_number(scale = 1e-3, suffix = "k", big.mark = "'")  
  )   +
    labs(title = "Heatmap of Demographic Characteristics by City", 
       subtitle = "Source: CMS 2022",
       x = "Demography Chracteristics", 
       y = "City", 
       fill = "Count") +
  theme_minimal() +
    theme(
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank())
```