---
title: "Data Cleaning"
author: "Claudia Bardales"
date: "`r Sys.Date()`"
output: word_document
---


```{r }

my_data <- read.csv("Medicare_Part_D_Prescribers_by_Provider_2022.csv")


```
#libraries
```{r }
library(tidyverse)
#install.packages("sf")
library(sf)
# install.packages("viridis")
# install.packages("maps")
#install.packages("tigris")

#heatmap
options(tigris_use_cache = TRUE)

library(sf)
library(viridis)
library(maps)
library(tigris)
```
I will work only with us data, so I will filter out the rest of the data.
Prscrbr_NPI
Tot_Drug_Cst (Total drug cost)- include processing fee and other admin fees
Prscrbr_Type (Prescriber Type)
Prscrbr_State_Abrvtn (State abbreviation)
Tot_Clms (Total claims)
Opioid_Tot_Clms (Total opioid claims)
Antbtc_Tot_Clms
Brnd_Tot_Clms 
Gnrc_Tot_Clms
Opioid_Tot_Suply

#Categorical Variables--> Beneficiary Profile
Beneficiary Age
Beneficiary Race
Beneficiary Sex
"Bene_Avg_Age"                  
[70] "Bene_Age_LT_65_Cnt"             "Bene_Age_65_74_Cnt"             "Bene_Age_75_84_Cnt"            
[73] "Bene_Age_GT_84_Cnt"             "Bene_Feml_Cnt"                  "Bene_Male_Cnt"                 
[76] "Bene_Race_Wht_Cnt"              "Bene_Race_Black_Cnt"            "Bene_Race_Api_Cnt"             
[79] "Bene_Race_Hspnc_Cnt"            "Bene_Race_Natind_Cnt"           "Bene_Race_Othr_Cnt"  

```{r }
head(my_data)
unique(my_data$Prscrbr_Cntry)
my_data_clean <- my_data[my_data$Prscrbr_Cntry == "US", ]
colnames(my_data)
```

```{r }
my_data_clean <- my_data_clean |>
  select(
    Prscrbr_NPI,
    Tot_Drug_Cst,
    Prscrbr_Type,
    Prscrbr_State_Abrvtn,
    Prscrbr_City,
    Prscrbr_zip5,
    Tot_Clms,
    Opioid_Tot_Clms,
    Antbtc_Tot_Clms,
    Brnd_Tot_Clms,
    Gnrc_Tot_Clms,
    Opioid_Tot_Suply,
    Bene_Age_LT_65_Cnt,
    Bene_Age_65_74_Cnt,
    Bene_Age_75_84_Cnt,
    Bene_Age_GT_84_Cnt,
    Bene_Feml_Cnt,
    Bene_Male_Cnt,
    Bene_Race_Wht_Cnt,
    Bene_Race_Black_Cnt,
    Bene_Race_Api_Cnt,
    Bene_Race_Hspnc_Cnt,
    Bene_Race_Natind_Cnt,
    Bene_Race_Othr_Cnt
  )

str(my_data_clean)
```



```{r }
#function to remove , and $
remove_dollar_comma<- function(x){
  as.numeric((gsub('[$,]',"",x)))
}

#apply to all columns except Prscrbr_NPI, Prscrbr_Type, Prscrbr_State_Abrvtn, Prscrbr_zip5
my_data_clean <- my_data_clean |>
  mutate(across(
    .cols = -c(Prscrbr_NPI, Prscrbr_Type, Prscrbr_State_Abrvtn,Prscrbr_City, Prscrbr_zip5), 
    .fns = ~remove_dollar_comma(.), 
    .names = "{.col}" 
  ))

str(my_data_clean)

```

#check unique values per columns

```{r }
#check na
na_summary <- colSums(is.na(my_data_clean))
na_summary

#filter no na (dropped 5 rows in total)
my_data_clean <- my_data_clean |>
  filter(!is.na(Prscrbr_zip5) & !is.na(Prscrbr_Type))

#replace na to 0
my_data_clean[is.na(my_data_clean)] <- 0

#check na
na_summary <- colSums(is.na(my_data_clean))
na_summary
```

```{r }
str(my_data_clean)
```
#Since there is 1332159 obs. I will reduce it to only state of florida (Florida total obs. = 89,231)
```{r }
my_data_clean_fl<-my_data_clean[my_data_clean$Prscrbr_State_Abrvtn=="FL",]
my_data_clean_fl
```
#convert zipcode to txt
```{r }
my_data_clean_fl$Prscrbr_zip5 <- as.character(my_data_clean_fl$Prscrbr_zip5)

```

#writefile as csv 
```{r }
write.csv(my_data_clean_fl, "my_data_clean_fl.csv", row.names = FALSE)

```


```{r }

```

